---
title: Tests
---

### MSSQL

```{bash}
echo 42
```

On Ubuntu, run this in the bash shell terminal on the system:

```{bash}
if ! [[ "18.04 20.04 22.04 24.04" == *"$(grep VERSION_ID /etc/os-release | cut -d '"' -f 2)"* ]];
then
    echo "Ubuntu $(grep VERSION_ID /etc/os-release | cut -d '"' -f 2) is not currently supported.";
    exit;
fi

# Download the package to configure the Microsoft repo
curl -sSL -O https://packages.microsoft.com/config/ubuntu/$(grep VERSION_ID /etc/os-release | cut -d '"' -f 2)/packages-microsoft-prod.deb
# Install the package
sudo dpkg -i packages-microsoft-prod.deb
# Delete the file
rm packages-microsoft-prod.deb

# Install the driver
sudo apt-get update
sudo ACCEPT_EULA=Y apt-get install -y msodbcsql18
# optional: for bcp and sqlcmd
sudo ACCEPT_EULA=Y apt-get install -y mssql-tools18
echo 'export PATH="$PATH:/opt/mssql-tools18/bin"' >> ~/.bashrc
source ~/.bashrc
# optional: for unixODBC development headers
sudo apt-get install -y unixodbc-dev
```

Then run this to verify this database connection type is now available.

```{bash}
odbcinst -q -d
```

### Test SQL Server to DuckDB Pipeline

From sql file

```{r}
     
sql <- readLines("scripts/SQL/msTest.sql")

# Execute the SQL commands
dbExecute(mssql_con, paste(sql, collapse = "\n"))

result <- tbl(mssql_con, I("dbo.inventory"))

print(result)
```

Or from sql code chunk

```{sql}
#| connection: mssql_con
USE TestDB;

CREATE TABLE dbo.Inventory
(id INT,
name NVARCHAR (50),
quantity INT,
PRIMARY KEY (id));

INSERT INTO dbo.Inventory
VALUES (1, 'banana', 150);

INSERT INTO dbo.Inventory
VALUES (2, 'orange', 154);
```

Lazy query on mssql database table

```{r}
#| label: lazyMssql
tbl(mssql_con, I("dbo.Inventory"))
```

Register mssql table with duckdb

```{r}
#| label: regMssql2ddb
tbl(mssql_con, I("dbo.Inventory")) |>
     duckdb_register(conn = duckdb_con, name = I("dbo.Inventory"))
```

return duckcdb registered mssql table as dataframe output

```{r}
#| label: dfOutRegMssqlddb
result <- dbGetQuery(duckdb_con, "SELECT * FROM 'dbo.Inventory';")
print(result)
```

```{r}
#| label: normalCalcs
result |>
     summarize(q_sum = sum(quantity))
```

or lazy query registered mssql table with duckdb

```{r}
#| label: regMssql2Ddb
tbl(duckdb_con, I("'dbo.Inventory'"))
```

do some lazy calculations

```{r}
#| label: lazyCalculations
tbl(duckdb_con, I("'dbo.Inventory'")) |>
     summarize(
     q_sum = sum(quantity), 
     q_avg = mean(quantity)
     )
```

if you want to output as an object

```{sql}
#| label: output2Robject
#| connection: duckdb_con
#| output.var: "summary_of_hourly"
SUMMARIZE 
SELECT * 
FROM hourly;
```

## Test Performance

### Compare

```{r}


gc()

# Profile the original code
profvis({
  result_original <- tbl(duckdb_con, "hourly") |>
    select(
      date, 
      precipitation_probability, 
      precipitation, 
      rain, 
      snowfall
    ) |>
    collect() |>
    mutate(
      time_only = format(date, "%H:%M:%S"),
      common_date = as.POSIXct(paste("1970-01-01", time_only), format = "%Y-%m-%d %H:%M:%S"),
      day = format(date, "%b %d")  # Extract day in "Feb 19" format
    )
})

```

```{r}
gc()

# Measure memory usage before running the code
mem_before_original <- mem_used()

# Run the original code
result_original <- tbl(duckdb_con, "hourly") |>
  select(
    date, 
    precipitation_probability, 
    precipitation, 
    rain, 
    snowfall
  ) |>
  collect() |>
  mutate(
    time_only = format(date, "%H:%M:%S"),
    common_date = as.POSIXct(paste("1970-01-01", time_only), format = "%Y-%m-%d %H:%M:%S"),
    day = format(date, "%b %d")  # Extract day in "Feb 19" format
  )

# Measure memory usage after running the code
mem_after_original <- mem_used()

# Calculate memory used by the original code
mem_used_original <- mem_after_original - mem_before_original

# Print memory usage
print(paste("Memory used by original code:", mem_used_original))
```

```{r}
gc()
original_time <- system.time({
  result_original <- tbl(duckdb_con, "hourly") |>
    select(
      date, 
      precipitation_probability, 
      precipitation, 
      rain, 
      snowfall
    ) |>
    collect() |>
    mutate(
      time_only = format(date, "%H:%M:%S"),
      common_date = as.POSIXct(paste("1970-01-01", time_only), format = "%Y-%m-%d %H:%M:%S"),
      day = format(date, "%b %d")  # Extract day in "Feb 19" format
    )
})

# Print the execution time
print("Original code execution time:")
print(original_time)

```

```{r}
gc()

# Profile the pure SQL approach
profvis({
  result_sql <- dbGetQuery(duckdb_con, "
    SELECT 
      date, 
      precipitation_probability, 
      precipitation, 
      rain, 
      snowfall,
      strftime(date, '%H:%M:%S') AS time_only,
      strptime('1970-01-01 ' || strftime(date, '%H:%M:%S'), '%Y-%m-%d %H:%M:%S') AS common_date,
      strftime(date, '%b %d') AS day
    FROM 
      hourly;
  ")
})
```

```{r}

gc()

# Measure memory usage before running the code
mem_before_sql <- mem_used()

# Run the pure SQL approach
result_sql <- dbGetQuery(duckdb_con, "
  SELECT 
    date, 
    precipitation_probability, 
    precipitation, 
    rain, 
    snowfall,
    strftime(date, '%H:%M:%S') AS time_only,
    strptime('1970-01-01 ' || strftime(date, '%H:%M:%S'), '%Y-%m-%d %H:%M:%S') AS common_date,
    strftime(date, '%b %d') AS day
  FROM 
    hourly;
")

# Measure memory usage after running the code
mem_after_sql <- mem_used()

# Calculate memory used by the pure SQL approach
mem_used_sql <- mem_after_sql - mem_before_sql

# Print memory usage
print(paste("Memory used by pure SQL approach:", mem_used_sql))
```

```{r}
gc()
# Measure the performance of the new code
new_time <- system.time({
  result_new <- dbGetQuery(duckdb_con, "
    SELECT 
      date, 
      precipitation_probability, 
      precipitation, 
      rain, 
      snowfall,
      strftime(date, '%H:%M:%S') AS time_only,
      strptime('1970-01-01 ' || strftime(date, '%H:%M:%S'), '%Y-%m-%d %H:%M:%S') AS common_date,
      strftime(date, '%b %d') AS day
    FROM 
      hourly;
  ")
})

# Print the execution time
print("New code execution time:")
print(new_time)
```

```{r}
gc()
# Profile the lazy query approach
profvis({
  hourly_lazy <- tbl(duckdb_con, "hourly")
  
  lazy_query <- hourly_lazy |>
    select(
      date, 
      precipitation_probability, 
      precipitation, 
      rain, 
      snowfall
    ) |>
    mutate(
      time_only = sql("strftime(date, '%H:%M:%S')"),  # Use SQL for DuckDB functions
      common_date = sql("strptime('1970-01-01 ' || strftime(date, '%H:%M:%S'), '%Y-%m-%d %H:%M:%S')"),
      day = sql("strftime(date, '%b %d')")  # Extract day in "Feb 19" format
    )
  
  result_lazy <- lazy_query |>
    collect()
})
```

```{r}

gc()

# Measure memory usage before running the code
mem_before_lazy <- mem_used()

# Run the lazy query approach
hourly_lazy <- tbl(duckdb_con, "hourly")

lazy_query <- hourly_lazy |>
  select(
    date, 
    precipitation_probability, 
    precipitation, 
    rain, 
    snowfall
  ) |>
  mutate(
    time_only = sql("strftime(date, '%H:%M:%S')"),  # Use SQL for DuckDB functions
    common_date = sql("strptime('1970-01-01 ' || strftime(date, '%H:%M:%S'), '%Y-%m-%d %H:%M:%S')"),
    day = sql("strftime(date, '%b %d')")  # Extract day in "Feb 19" format
  )

result_lazy <- lazy_query |>
  collect()

# Measure memory usage after running the code
mem_after_lazy <- mem_used()

# Calculate memory used by the lazy query approach
mem_used_lazy <- mem_after_lazy - mem_before_lazy

# Print memory usage
print(paste("Memory used by lazy query approach:", mem_used_lazy))
```

```{r}
gc()
# Measure the performance of the lazy query approach
lazy_time <- system.time({
  hourly_lazy <- tbl(duckdb_con, "hourly")
  
  lazy_query <- hourly_lazy |>
    select(
      date, 
      precipitation_probability, 
      precipitation, 
      rain, 
      snowfall
    ) |>
    mutate(
      time_only = sql("strftime(date, '%H:%M:%S')"),  # Use SQL for DuckDB functions
      common_date = sql("strptime('1970-01-01 ' || strftime(date, '%H:%M:%S'), '%Y-%m-%d %H:%M:%S')"),
      day = sql("strftime(date, '%b %d')")  # Extract day in "Feb 19" format
    )
  
  result_lazy <- lazy_query |>
    collect()
})

# Print the execution time
print("Lazy query code execution time:")
print(lazy_time)
```

```{r}
print(paste("Memory used by original code:", mem_used_original))
print(paste("Memory used by pure SQL approach:", mem_used_sql))
print(paste("Memory used by lazy query approach:", mem_used_lazy))
```

```{r}
print("Original code execution time:")
print(original_time)

print("Pure SQL code execution time:")
print(new_time)

print("Lazy query code execution time:")
print(lazy_time)
```

## Output

```{r}
test <- tbl(duckdb_con, "minutely") |>
select("date", "temperature_2m") |>
collect() |>
mutate(
time_only = format(date, "%H:%M:%S"),
common_date = as.POSIXct(paste("1970-01-01", time_only), format = "%Y-%m-%d %H:%M:%S"),
day = format(date, "%b %d")  # Extract day in "Feb 19" format
)

test |>
ggplot(mapping = aes(x = common_date, y = temperature_2m)) +
  geom_line() +
  labs(title = "Temperature Trends", x = "Date", y = "Temperature (°F)") +
  theme_minimal() +
   scale_x_datetime(
     labels = label_time("%l %p"),
     breaks = "7 day",
     minor_breaks = "1 hour",
     guide = guide_axis(n.dodge = 2)
     ) +
   facet_grid(.~ day)  # One facet per day
```

```{python}
import requests
import pandas as pd
from bs4 import BeautifulSoup
 
# link for extract html data
 
 
def getdata(url):
    r = requests.get(url)
    return r.text
 
 
htmldata = getdata("https://www.goodreturns.in/petrol-price.html")
soup = BeautifulSoup(htmldata, 'html.parser')
 
# Declare string var
# Declare list
mydatastr = ''
result = []
 
# searching all tr in the html data
# storing as a string
for table in soup.find_all('tr'):
    mydatastr += table.get_text()
 
# set according to your required
mydatastr = mydatastr[1:]
itemlist = mydatastr.split("\n\n")
 
for item in itemlist[:-5]:
    result.append(item.split("\n"))
 
# Calling DataFrame constructor on list
df = pd.DataFrame(result[:-8])
df

```

## Coordinate Examples

| Location Name | Address | Lat, Long |
|------------------------|------------------------|------------------------|
| \- | Intersection of Hwy 70 & 270 | 38.748, -90.439 |
| The St. Louis Regional Freightway | \- | 38.627480, -90.189781 |
| Port of St. Louis | \- | 38.609215, -90.204275 |
| USPS Office Kansas City International Airport | \- | 39.304413, -94.720711 |
| Walmart Distribution Center | 1100 Matlock Dr, Saint James, MO 65559-9026 | 38.002338, -91.634422 |
| KCS International Freight Gateway | 3301 East 147th Street, Kansas City, MO 64147 | 38.84930037041913, -94.55499143307868 |

: Real locations that might be points of interest for truckers.
